```{r setup} #can be r or r setup, r setup will run our library when even we need it}
install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

We will be practicing with `dplyr` functions using the `starwars` dataset

#to get that block that's highlighting dplyr and starwars, highlight it and click the code button by the bold and italicize key

```{r}
starwars
head(starwars)
```

## Subset rows with `filter`

filter() allows us to subset the rows of a dataset based on the values

```{r}
##filter for rows with human Star Wars characters
filter(starwars, species == "Human")

##save to new r object
humans <- filter(starwars, species == "Human")
humans
filter(humans, sex == "male")

##filter multiplue colums at once
filter(starwars, species == "Human", sex == "male")

#
```

`!` is "not"

\_ is equal

`&` is "and"

`|` is "or"

`<` is "less than"

`>`is "greater than"

`is.no()` returns TRUE if value is NA

```{r}
## return charaters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
##how to get all characters that are NOT human
filter(starwars, species != "Human")
```

variable_name %in% vector_of_values return TRUE if the variable_name is equal to one of the values in vector_of_values

```{r}
skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan")#returns false bc "grey" is not in vector of colors


hair_color %in% c("brown", "blond")

filter(starwars, hair_color %in% c("blond", "brown") )
```

grepl("characters_of_Interest", variable_name) returns TRUE if the variable_name contains the characters_of_interest.

```{r}
filter(starwars, grepl("blond", hair_color))
filter(starwars, grepl("blue", eye_color))
```

You can use `<`, `>`, `<=`, `>=` to compare numeric variables or categorical variables

-   nominal categorical variable (character vectors) are ranked alphabetically

-   ordinal categorical variable (factors) are ranked based on the built-in order

    ```{r}
    ##starwars characters with a name starting with letters before "D"
    filter(starwars, name < "D")

    str(diamonds)
    filter(diamonds, cut > "Good") #ranking is based on levels of 

    "D" < "d"
    "d" < "D"
    ```

between(x, left, right) is a shorthand for x \>= left & x \<= right

```{r}
between(5, 1, 7) #true because 5 is less tnan or equal to 7 and greater than or equal to 1
filter(starwars, between(mass, 50, 100))
```

### Practice

Starwars

filter(starwars, is.na(mass))

```{r}
#1.are missing a hair color
filter(starwars, is.na(hair_color))

#2. species that are shorter than 100cm and are not droids
filter(starwars, height < 100 & species != "Droid")

#3. born between the years of 100 and 200
filter(starwars, between(birth_year, 100, 200))

#my own choice
filter(starwars, is.na(homeworld)) #species with an unknown homeland


```

## Order rows with `arrange`

arrange() reorders the rows based on the values of column(s)

```{r}
arrange(starwars, name) #names in alphabetical order
#order of character values: a AbBcCdD

##order by species and birth year
arrange(starwars, species, birth_year)

arrange(starwars, desc(birth_year))#desc is descending, we use it to get the value in descending order


#practice

#1. arrage the starwars dataset from tallest to shortest characters

arrange(starwars, desc(height))

#2. arrage shortest to tallest by species

arrange(starwars, height)

#3. alphabetize the charatcers by their hair color then eye color

arrange(starwars, hair_color, eye_color)

#4. 

arrange(starwars, species)

```

## Subbset columns with `select()`

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```

we can also select multiple columns in a row

```{r}
#all colums between name and eye_color(inclusive)
select(starwars, name:eye_color)
```

we can also remove colums by a name using a

```{r}
starwars
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

helpful functions

everything() is a shortcut for "everything else"

```{r}
select(starwars, "species", everything())
starwars_names = sort(names(starwars))
starwars_names
select(starwars, starwars_names)
```

`stars_with("abc")` matches column names that start with "abc"

`ends_with("abc")` matches column names that end with "abc"

`contains("abc")` matches column names that contain ("abc")

```{r}
select(starwars, starts_with("h"))

select(starwars, contains("or"))


#practice

select(starwars, contains(".."))
select(starwars, contains("color"), birth_year)
select(starwars, hair_color:birth_year)
select(starwars)
select(starwars, hair_color, skin_color, birth_year)


```

to tell us which species in common to each homeworld and if there are outlines or uncommon species in a certain homeworld.

Add new variable with `mutate`

```{r}
starwars_small = select(starwars, height, mass, species)
starwars_small

mutate(starwars_small, 
       height_m = height/100)

```

we can also create multiple variables at once and we can refer to the variables we just created

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

#only keep new variable
transmute(starwars_small,
          height_m = height/100,
          bmi = mass/height_m^2)
```

summary statisics give one value for the entire column: `sum(), mean()`

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = T))

mutate(starwars_small,
       row_addition = height + mass,
       summary_sum = sum(height, mass, na.rm = T))
```

group by a variable using .by parameter

```{r}
mutate(starwars_small,
       species_mass = mean(mass, na.rm = T),
       .by = species)

mutate(starwars_small,
       weight = mass*2.2)

```

change by the type of variable

```{r}
mutate(starwars_small,
       mass = as.logical(mass)) #if the number is zero, its false, greater or equal to 1 us true
```

### summarize data with `summarize`

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T))

#group by species
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          .by = species)
#the only colums kept are the colums you group by and the colums you create
```

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = species)

summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = sex, species)

```
